Synthesizing Symmetric Lenses

### 2018-05-31 papers/SymmetricLenses/main.tex
clean up overview

  We develop a type-directed synthesis algorithm for generating lenses in this
  language from regular expressions describing the data formats. A significant
  technical challenge in this work arises from the fact that our target language
  permits synchronization between formats that don't have exactly the same data
  -- data in one format need not be present in the other format. This gives the
  synthesis algorithm much more freedom, as data can be completely
  unsynchronized (edits in one format are never propagated), completely
  synchronized (the formats are in bijective correspondence), or anywhere in
  between. Under the assumption that users typically want to synchronize as much
  data as possible, we provide an information theoretic metric for how much data
  is left unsynchronized based on the theory of \emph{stochastic regular
    expressions}. Furthermore, we use this metric to develop a greedy synthesis
  algorithm that tries to minimize unsynchronized data. We experimentally
  demonstrate that our algorithm can efficiently synthesize simple symmetric
  lenses on a variety of benchmark suites consisting of synchronizing file
  formats from configuration files, application-specific data storage files, and
  data cleaning tasks.

@@

\emph{Symmetric lenses} formalize such conversion functions. However, these
symmetric lenses use external state for synchronization, in the form of a
\emph{complement}. This makes them difficult to use in real-world
synchronization tasks, as users must maintain the synchronized files synchronize
as well as the complements. The complement further complicates synthesis, as
input-output examples are no longer sufficient, instead edit sequences must be
provided.

@@

A primary difficulty is that there are a relatively many well-typed simple
symmetric lenses. As symmetric lenses permit data to be present in only one
format, for any data component there is now a choice: is this data only relevant
to this format and as such should be left unsynchronized, or is this data
present in both formats and as such should be synchronized.

Operate under the assumption that users wish to synchronize as much data as
possible, our algorithm try to synchronize as much data as the types and
examples allow. This task gets harder when there are more complex decisions
involved. For example, if I have two fields in one format, and one in the other,
which field on the left should be synchronized to the one of the right? Or would
it be better for the field on the right to be synchronized to a mish-mash of
subcomponents from each field? To answer these questions, we develop a cost
metric for reasoning about which lens synchronizes the most information.

@@

Our model for this is based on \emph{stochastic regular expressions}~\cite{?}. A
probability distribution for strings is represented by a stochastic regular
expression, and we develop a way to syntactically infer the expected information
content (or \emph{entropy}) of an unambiguous regular expression. This entropy
is used to calculate the expected information content to recover a string, when
given a string it is synchronized to.

@@

We develop an algorithm that synthesizes a low-cost lens, shown in
Figure~\ref{fig:high-level-algorithm}. This algorithm consists of two
communicating synthesizers, one that proposes candidate stochastic regular
expression pairs that it thinks will align well, and one that greedily searches
for a lens between those two regular expressions that minimizes cost and
involves no rewrites. When the first algorithm thinks it has found the best
possible lens, it returns that lens.

Our first algorithm, \RXSearch, proposes candidate stochastic regular
expressions. These candidate stochastic regular expressions are generated by
traversing \emph{star semiring} equivalences on stochastic regular expressions.
These proposed stochastic regular expression pairs have equivalent languages and
probability distributions as the original ones: we prove the star-semiring
rewrites this synthesizer applies are sound rewrites on stochastic regular
expression.

Our second algorithm, \GreedySynth, utilizes alternative languages for
stochastic regular expressions and lenses, \emph{stochastic DNF regular
  expressions} and \emph{symmetric DNF lenses}. While these languages are hard
to read and understand, they permit an efficient type-directed synthesis
algorithm. We develop our cost metric on DNF lenses, and provide a greedy
algorithm that tries to find a symmetric DNF lens of minimal cost. The
synthesized symmetric DNF lens is then converted to a lens in the original
simple symmetric lens language.

We evaluate our algorithm on a benchmark suite consisting of the original
Optician benchmarks, as well as additional benchmarks consisting of
synchronization tasks where the data is not in bijective correspondence.

In summary, our contributions are:
\begin{enumerate}
\item We develop \emph{simple symmetric lenses} a new formulation for
  bidirectional programs that can project data from both sides~(\S\ref{sec:overview}),
  and describe a language that expresses such lenses on string
  data~(\S\ref{sec:ssl}).
\item We develop a means to traverse stochastic regular expression equivalences,
  while preserving probability distributions and
  languages~(\S\ref{sec:stoch-rx}).
\item We develop stochastic DNF regular expressions and DNF lenses. We develop a
  sound algorithm for converting stochastic regular expressions to stochastic
  DNF regular expressions~(\S\ref{sec:dnf-regex-lens}), and a
  information-theoretic cost metric on DNF
  lenses~(\S\ref{sec:symmetric-dnf-lenses}).
\item We extended Boomerang with our simple symmetric lens combinators. We
  updated the Optician synthesis system to synthesize symmetric lenses with our
  new algorithm.
\item We evaluate our synthesis algorithm on Opticians benchmark suite as well
  as new, symmetric benchmarks, and find that we lose very little synthesis
  time, while being significantly more expressive (\S\ref{sec:evaluation}).
\end{enumerate}
% end introduction

@@

We motivate our problem with an extended example. Consider the previously
described example of keeping Linux and Windows scheduled jobs synchronized. In
this example, we describe keeping synchronized a crontab (the file containing
cron jobs) and a Scheduled Tasks file, scheduled.job. These are each formats
specifying when to run recurring tasks. For simplicity of description, we use a
minified version of each of these, with examples and regular expressions shown
in Figure~\ref{fig:minimized-representations}.

For task is set in these files at specific times each hour. In both these
formats, users can specify specific minutes they want to run the program during,
or they can specify they want to run the program at every minute.

In this example, we wish for the jobs to be synchronized across our windows and
linux devices. However, the commands that need to be run are typically different
for each device. Paths are different in Windows and Linux, and while there is
typically a Windows equvialent for Linux commands, that equivalent command is
not always named the same thing.  So, while we want to be able to add commands
and edit how often they are run, we don't want to synchronize which commands are
run themselves.

To synchronize these files, a few functions are needed. Two \emph{creates} are
needed; if a crontab is initialized, a synchronized Scheduled Tasks file is also
created, and vice versa. Furthermore, \emph{puts} are needed, functions that
take an updated crontab and an old Scheduled Tasks file, and output a new
Scheduled Tasks file with the information synchronized to the updated crontab.
To have a simple symmetric lens is to have these required synchronization
functions. Formally, a simple symmetric lens between $X$ and $Y$ comprises 4
functions:
\begin{enumerate}
\item $\CreateR \OfType X \rightarrow Y$
\item $\CreateL \OfType Y \rightarrow X$
\item $\PutR \OfType X \rightarrow Y \rightarrow Y$
\item $\PutL \OfType Y \rightarrow X \rightarrow X$
\end{enumerate}
Furthermore, if a file is not edited, but is then resynchronized using a put
function, no data should be changed. To be a lens, one must satisfy the
following ``round-tripping'' laws that guarantee identity edits won't change the
file data.
\begin{gather}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{gather}

@@

\subsection{Programming Symmetric Lenses} We now build up a symmetric lens by
example. Our symmetric lens expressions are typed between regular expression
pairs, if \lstinline{$\Lens$ : $\Regex \Leftrightarrow \RegexAlt$}, then $\Lens$
encodes a lens between $\LanguageOf{\Regex}$ and $\LanguageOf{\RegexAlt}$. At
the end, we want our symmetric lenses to by typed between \lstinline{crontab}
and \lstinline{scheduled_job}.

The identity lens takes, as argument, a regular expression, and performs the
identity function on strings in that regular expression. The identity lens fully
synchronizes the data formats: when one data format is updated, the other format
gets the exact same value. In our example, when a job is told to run at a
specific time in crontab, it is also expected to run at that same time in the
Scheduled Jobs list, so the identity lens would propagate that information well.
%
\begin{lstlisting}
id(minute) : minute $\Leftrightarrow$ minute
\end{lstlisting}
%

Where the identity lens synchronizes data, the disconnect lens does the
opposite: it leaves data unsynchronized. The disconnect lens takes 4 arguments:
two regular expressions and two strings. On creates, because there is no
association between data in the first format and data in the second format,
designated default strings $\String$ and $\StringAlt$ are put into the other
format. In our extended example, the Linux commands should not by synchronized
with the Windows commands, so a disconnect lens would be used. We want our
default behavior to be failures, under the assumption this failure will raise a
warning, in case somebody forgets to find the analogous command of the other
operating system.
%
\begin{lstlisting}
disc(command,command_xml,"false","<Command>SET ERRORLEVEL=1</Command>") 
  : command $\Leftrightarrow$ command_xml
\end{lstlisting}
%
When building a disconnect lens between two strings, one of which is merely the
empty string, the lenses \lstinline{ins} and \lstinline{del} can instead be
used, where \lstinline{ins($t$) = disc("",$t$,"",$t$)} and
\lstinline{del($s$) = disc($s$,"",$s$,"")}.

Finally, there are a number of lenses that compositionally build up new lenses
from old ones. For example, data can appear in one form or another, a program
can run during a specific set of minutes, or that program can run the every
minute. If \lstinline{minutes_map} and \lstinline{alltimes_map} are lenses that
synchronize those settings in each format, then those lenses can be composed
with an \OrLens lens.
%
\begin{lstlisting}
or(minutes_map,alltimes_map) : times $\Leftrightarrow$ xml_time
\end{lstlisting}

Data can also be concatenated, and so lens mappings can be concatenated.
However, data concatenations can only happen in a number of ways, as data can
appear in the same order, and data can be reordered. For example, in the
crontab, time data appears before command data, whereas in the Scheduled Jobs
file, command data comes before time data. If \lstinline{command_map} provides a
lens between commands in each format, and \lstinline{time_map} provides a lens
between time data in each format, those lenses can be composed with a \SwapLens
lens.
%
\begin{lstlisting}
swap(command_map,time_map) : line $\Leftrightarrow$ xml_task
\end{lstlisting}
%
Two lenses, $\Lens_1$ and $\Lens_2$, can be concatenated as
\lstinline{concat($\Lens_1$,$\Lens_2$)} or as the infix style
\lstinline{$\Lens_1$ . $\Lens_2$} for data that is not reordered.

Lastly, the \IterateLens lens synchronizes repeated data, synchronizing each
element of the iteration, and synchronizing the number of iterations. For
example, if \lstinline{full_task_map} synchronizes crontab lines prepended by
two newlines and xml for a single task, iterating that lens builds a lens that
synchronizes a list of tasks.
%
\begin{lstlisting}
iterate(full_task_map) : ("\n\n" . line)$^*$ $\Leftrightarrow$ xml_task$^*$
\end{lstlisting}
%

@@

let task_map = swap(command_map,time_map) in
let full_task_map = 
  ins("<Task>") . del("\n\n") . task_map . ins("</Task>") 
in
let repeated_task_map = iterate(full_task_map)
let nonempty_task_map = ins("<Tasks>") . repeated_task_map . ins("</Tasks>")

@@

Using these combinators, we can build up a lens that synchronizes crontabs and
Scheduled Task files, shown in Figure~\ref{fig:example_lens}. This lens uses one
typing rule not previously mentioned: type equivalence.  While
\lstinline{scheduled_job} is a equivalent to a lens with an outermost
disjunction, it is not syntactically equivalent to one.  Without type
equivalence, the provided lens wouldn't actually be well-typed, but type
equivalence states that if a lens is typed between a pair of regular
expressions, it is also typed between an equivalent pair.

Note that, while the programmed lens should be quite simple, it was relatively
difficult to write. We had to provide a lot of scaffolding to map between the
desired formats, even though the idea was quite simple: synchronize the times
but leave the commands unsynchronized. Furthermore, this is inherent to manually
writing these synchronization functions, as a programmer must describe which
components are synchronized with which, and which components are scaffolding.

\subsection{Synthesizing Symmetric Lenses}
To alleviate these difficulties, we wish to synthesize symmetric lenses from
types and examples. Given a pair of regular expressions and a set of
input-output examples, we want to synthesize a symmetric lens expression that is
typed between the regular expression pair, and satisfies the input/output
examples. For example, we wish to synthesize a lens from \lstinline{crontab} and
\lstinline{scheduled_job}, using the strings in
Figure~\ref{fig:minimized-representations} as input-output example for \CreateR.

However, our simple symmetric lens combinators permit many well-typed programs
between two regular expressions. For example, one possible lens between
\lstinline{crontab} and \lstinline{scheduled_job} is the one in
Figure~\ref{fig:example_lens}, but the lens
%
\begin{lstlisting}
disc(crontab,scheduled_job,rep(crontab),rep(scheduled_job))
\end{lstlisting}
%
is also a valid lens (where \lstinline{rep($\Regex$)} is a string in the
language of $\Regex$). Furthermore, adding the synchronized files from
Figure~\ref{fig:minimized-representations} as an input-output example helps
little, but the default strings given to \Disconnect merely need to be those
strings and the lens still satisfies the examples. While the disconnect lens is
prohibited by two examples, or by a carefully chosen input-output example on
\PutR or \PutL, this problem continues presenting itself for the sublenses, and
is highlighted even more on larger formats. While we could merely ask
programmers to provide large, tight specifications for their synthesis tasks, we
would prefer the algorithm to choose the right lens automatically. Instead of
merely finding \emph{any} satisfying lens, we wish to synthesize the \emph{best}
satisfying lens.

But what is the best satisfying lens?  We think the best satisfying lens is the
one that synchronizes as much data as is permitted by the type system and the
examples.  More formally, we wish to minimize the expected number of bits
required to recover data in one format from the other.  The expected number of
bits is a well known concept in information theory, the expected number of bits
of a data source is the \emph{entropy} of that data source.

The entropy of a data source is based on the probability distribution of that
data source. Given a set $S$, and a probability distribution over that set $P
\OfType S \rightarrow \Reals$, the entropy of $(S,P)$ is $-\Sigma_{s\in
  S}P(s)\Log_2P(s)$. While we have ways of expressing sets, we do not have
probability distributions. In this work, we use \emph{stochastic regular
  expressions}, regular expressions annotated with probabilities that express
languages and probability distributions over those languages. In this work, we
guess a reasonable probability distribution from the syntax, though if users
manually provided stochastic regular expressions, or large suite of data in the
formats through which we could learn the annotations, we could instead use that.

With sets and probability distributions, we can find the entropy of the
stochastic regular expression, given another stochastic regular expression, and
a lens that shows which strings are synchronized with which. We aim to find the
lens that minimizes that entropy.  Now that we have a notion of best, we need a
way to search for that best lens.

Prior work on lens synthesis works with two communicating synthesizers, one that
proposes regular expressions, semantically but not syntactically equivalent to
each other, and another that finds lenses between them, if there are any lenses
between them that don't use a type equivalence rule involving star equivalences.
We utilize this same framework, with two modifications. (1) The first
synthesizer does not propose regular expressions, but stochastic regular
expressions, equivalent to the originals. We prove that these proposed
stochastic regular expressions don't merely have the same languages as the
originals, but also the same probability distributions. (2) Instead of having
the second synthesizer find \emph{any} lens between the types, \GreedySynth
instead uses a greedy algorithm to try finding the \emph{best} lens between the
types. Furthermore, the first synthesizer proposes stochastic regular
expressions with equivalent languages and probability distributions, instead of
just equivalent languages.

This poses a new problem: unless we find a lens with no cost, further search may
find a better lens. While we may have found a great lens between two regular
expressions, a different regular expression pair may provide an even better
lens. There may be a great lens between \lstinline{crontab} and
\lstinline{scheduled_job} that uses no star equivalences, but the lens shown in
Figure~\ref{fig:example_lens} does. However, there is a tension, we want to find
a good lens, but we also want to give quick feedback to the user. After finding
the lens in Figure~\ref{fig:example_lens}, the algorithm shouldn't spend much
longer searching for better lenses.

We address this by keeping running track of the best lens found thus far. Before
proposing new types, the regular expression generator, \RXSearch, evaluates
whether it should continue proposing regular expression pairs, or whether it is
content with the current best. For a given regular expression pair, \RXSearch
looks at the number of rewrite rules performed on the pair, and identifies how
many pairs with involving the same number of rewrites may be proposed. If there
are a large number of pairs that can be proposed with that many rewrites, or the
current lens is very well synchronized, \RXSearch does not propose that pairs.
In this way, \RXSearch will continue searching for better lenses, but does not
tackle huge search spaces unless it needs to. In this way, the algorithm can
provide lenses quickly to the users, which users can refine with additional
examples.
% end overview

@@

\subsection{\GreedySynth}
The second synthesizer, \GreedySynth, actually generates the lenses. It does
this by first converting the regular expressions into \emph{stochastic DNF
  regular expressions}, intuitively stochastic regular expressions with all
disjunctions distributed over all concatenations. Next, a good DNF lens is
greedily found between these stochastic DNF regular expressions, using the
probability annotations on the DNF regular expressions to find the cost.
Finally, this DNF lens is converted into a simple symmetric lens.

>>>>

  We develop a type-directed synthesis algorithm for generating lenses in this
  language from regular expressions describing the data formats. A significant
  technical challenge in this work arises from the fact that our target language
  permits synchronization between formats that don't have exactly the same data
  -- data in one format need not be present in the other format. This gives the
  synthesis algorithm much more freedom, as data can be completely unaligned
  (edits in one format are never propagated), completely aligned (the formats
  are in bijective correspondence), or anywhere in between. Under the assumption
  that users typically want to align as much data as possible, we provide an
  information theoretic metric for how much data is left unaligned based on the
  theory of \emph{stochastic regular expressions}. Furthermore, we use this
  metric to develop a greedy synthesis algorithm that tries to minimize
  unaligned data. We experimentally demonstrate that our algorithm can
  efficiently synthesize simple symmetric lenses on a variety of benchmark
  suites consisting of synchronizing file formats from configuration files,
  application-specific data storage files, and data cleaning tasks.

@@

\emph{Symmetric lenses} formalize such conversion functions. However, these
symmetric lenses use external state for synchronization, in the form of a
\emph{complement}. These complements makes symmetric lenses difficult to use in
real-world synchronization tasks, as users must maintain the synchronized files
as well as the complements. The complement further complicates synthesis, as
input-output examples are no longer sufficient, instead edit sequences must be
provided.

@@

A primary difficulty is that there are a relatively many well-typed simple
symmetric lenses. As symmetric lenses permit data to be present in only one
format, for any data component there is now a choice: is this data only relevant
to this format and as such should be left unaligned (changes to this data are
not propagated to the other format), or is this data present in both formats and
as such should be aligned (changes to this data change are propagated to the
other format).

Operating under the assumption that users wish to align as much data as
possible, our algorithm tries to align as much data as the types and examples
allow. This task gets harder when there are more complex decisions involved. For
example, if I have two fields in one format, and one in the other, which field
in the first format should be aligned to the sole field of the second format? Or
would it be better for the field on the right to be aligneded to a mish-mash
of subcomponents from each field? To answer these questions, we develop a cost
metric for reasoning about which lens aligns the most information.

@@

Our model for this is based on \emph{stochastic regular expressions}~\cite{?}.
Both a language and a probability distribution over that language is represented
by a stochastic regular expression, and we develop a way to syntactically infer
the expected information content (or \emph{entropy}) of an unambiguous regular
expression. This entropy is used to calculate the expected information content
to recover a string, when given a string it is synchronized to.

@@

We develop an algorithm that synthesizes a low-cost lens, shown in
Figure~\ref{fig:high-level-algorithm}. This algorithm consists of two
communicating synthesizers, one that proposes candidate stochastic regular
expression pairs that it thinks will align well, and one that greedily searches
for a lens between those two regular expressions that minimizes cost and does
not traverse star-related equivalences. When the first algorithm is content with
the lens it has found, it returns that lens.

Our first algorithm, \RXSearch, proposes candidate stochastic regular
expressions. These candidate stochastic regular expressions are generated by
traversing \emph{star semiring} equivalences on stochastic regular expressions.
These proposed stochastic regular expression pairs have equivalent languages and
probability distributions as the original ones: we prove the star-semiring
rewrites this synthesizer applies are sound rewrites on stochastic regular
expressions.

Our second algorithm, \GreedySynth, utilizes alternative languages for
stochastic regular expressions and lenses, \emph{stochastic DNF regular
  expressions} and \emph{symmetric DNF lenses}. While these languages are hard
to read and understand, they are highly normalized and permit an efficient
type-directed synthesis algorithm. We develop our cost metric on DNF lenses, and
provide a greedy algorithm that tries to find a symmetric DNF lens of minimal
cost. The synthesized symmetric DNF lens is then converted to a lens in the
original simple symmetric lens language.

We evaluate our algorithm on a benchmark suite consisting of synchronizing file
formats from configuration files, application-specific data storage files, and
data cleaning tasks. We find that our algorithm is able to synthesize all these
lenses in under 20 seconds.

In summary, our contributions are:
\begin{enumerate}
\item We develop \emph{simple symmetric lenses} a new formulation for
  bidirectional programs that can project data from both sides~(\S\ref{sec:overview}),
  and describe a language that expresses such lenses on string
  data~(\S\ref{sec:ssl}).
\item We develop a means to traverse stochastic regular expression equivalences,
  while preserving probability distributions and
  languages~(\S\ref{subsec:stoch-rx}).
\item We develop stochastic DNF regular expressions and DNF
  lenses~(\S\ref{subsec:greedy-synth}). We develop a sound algorithm for
  converting stochastic regular expressions to stochastic DNF regular
  expressions and a information-theoretic cost metric on DNF lenses.
\item We extended Boomerang with our simple symmetric lens combinators, and
  integrated our algorithm with the Optician synthesis system within Boomerang,
  allowing synthesis tasks as Boomerang
  expressions~(\S\ref{sec:implementation}).
\item We evaluate our algorithm on a benchmark suite consisting of synchronizing
  file formats from configuration files, application-specific data storage
  files, and data cleaning tasks (\S\ref{sec:evaluation}).
\end{enumerate}
% end introduction

@@

We motivate our problem with an extended example. Consider the previously
described example of keeping Linux and Windows scheduled jobs synchronized. In
this example, we describe keeping the file containing cron jobs (crontab) and a
Scheduled Tasks file (scheduled.job) synchronized. These are each formats
specifying when to run recurring tasks. For simplicity of description, we use a
minified version of each of these, with examples and regular expressions shown
in Figure~\ref{fig:minimized-representations}. Each file contains a list of
tasks set to run at specific times every hour. In both these formats, users can
either provide specific minutes they want to run the program during or they can
specify they want to run the program at every minute.

We wish for the jobs to be synchronized across our Windows and Linux devices.
However, the commands that need to be run are typically different for each
device. Paths are different in Windows and Linux, and while there is typically a
Windows equvialent for Linux commands, that equivalent command is not always
named the same thing. So while we want to be able to have the same number of
commands, with the commands running at the same times, we don't want to
align the actual commands that are run.

To synchronize these files, a few functions are needed. Two \emph{creates} are
needed; if a crontab is initialized, a synchronized Scheduled Tasks file should
also be created, and vice versa. Furthermore, \emph{puts} are needed; if a
crontab is updated, we should be able to combine it with a non-updated Scheduled
Tasks file, and output a new Scheduled Tasks file, synchronized to the updated
crontab. To have a simple symmetric lens is to have these required
synchronization functions. Formally, a simple symmetric lens between $X$ and $Y$
comprises 4 functions:
\begin{enumerate}
\item $\CreateR \OfType X \rightarrow Y$
\item $\CreateL \OfType Y \rightarrow X$
\item $\PutR \OfType X \rightarrow Y \rightarrow Y$
\item $\PutL \OfType Y \rightarrow X \rightarrow X$
\end{enumerate}
Furthermore, if a file is not edited, but is then resynchronized using a put
function, no data should be changed. To be a simple symmetric lens, one must
satisfy the following ``round-tripping'' laws, guaranteeing identity edits won't
change the file data.
\begin{gather}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{gather}

@@

\subsection{Programming Symmetric Lenses} We now build up a symmetric lens by
example. Our symmetric lens expressions are typed between regular expression
pairs, if \lstinline{$\Lens$ : $\Regex \Leftrightarrow \RegexAlt$}, then $\Lens$
encodes a simple symmetric lens between $\LanguageOf{\Regex}$ and
$\LanguageOf{\RegexAlt}$. At the end, we want our symmetric lenses to by typed
between \lstinline{crontab} and \lstinline{scheduled_job}.

The identity lens takes, as argument, a regular expression, and performs the
identity function on strings in that regular expression. The identity lens fully
aligns the data formats: when one data format is updated, the other format gets
the exact same value. In our example, when a job is told to run at a specific
time in crontab, it is also expected to run at that same time in the Scheduled
Jobs list. The identity lens propagates that information.
%
\begin{lstlisting}
id(minute) : minute $\Leftrightarrow$ minute
\end{lstlisting}
%

Where the identity lens aligns data, the disconnect lens does the opposite: it
leaves data unaligned. The disconnect lens takes 4 arguments: two regular
expressions and two strings. On creates, because there is no association between
data in the first format and data in the second format, designated default
strings are used for the other format. In our extended example, the Linux
commands should not be synchronized with the Windows commands, so a disconnect
lens would be used. We want our default behavior to be failures, under the
assumption this failure will raise a warning, in case somebody forgets to find
the analogous command of the other operating system.
%
\begin{lstlisting}
disc(command,command_xml,"false","<Command>SET ERRORLEVEL=1</Command>") 
  : command $\Leftrightarrow$ command_xml
\end{lstlisting}
%
When building a disconnect lens between two strings, one of which is merely the
empty string, the lenses \lstinline{ins} and \lstinline{del} can instead be
used, where \lstinline{ins($t$) = disc("",$t$,"",$t$)} and
\lstinline{del($s$) = disc($s$,"",$s$,"")}.

Finally, there are a number of lenses that compositionally build up new lenses
from old ones. For example, data can appear in one form or another: a program
can run during a specific set of minutes, or that program can run the every
minute. If \lstinline{minutes_map} and \lstinline{alltimes_map} are lenses that
synchronize those settings in each format, then those lenses can be composed
with an \OrLens lens.
%
\begin{lstlisting}
or(minutes_map,alltimes_map) : times $\Leftrightarrow$ xml_time
\end{lstlisting}

Data can also be concatenated, and so lens mappings can be concatenated.
However, synchronizing concatenated data can happen in a number of ways, as data
can appear in the same order, and data can be reordered. For example, in the
crontab, time data appears before command data, whereas in the Scheduled Jobs
file, command data comes before time data. If \lstinline{command_map} provides a
lens between commands in each format, and \lstinline{time_map} provides a lens
between time data in each format, those lenses can be composed with a \SwapLens
lens.
%
\begin{lstlisting}
swap(command_map,time_map) : line $\Leftrightarrow$ xml_task
\end{lstlisting}
%
Two lenses, $\Lens_1$ and $\Lens_2$, can be concatenated as
\lstinline{concat($\Lens_1$,$\Lens_2$)} or as the infix style
\lstinline{$\Lens_1$ . $\Lens_2$} for data that is not reordered.

Lastly, the \IterateLens lens synchronizes repeated data, aligning the number of
times the data is repeated. The \IterateLens combinator takes a lens as an
argument and synchronizes each element of the iteration with that lens. For
example, if \lstinline{full_task_map} synchronizes crontab lines prepended by
two newlines and xml for a single task, iterating that lens builds a lens that
synchronizes a list of tasks.
%
\begin{lstlisting}
iterate(full_task_map) : ("\n\n" . line)$^*$ $\Leftrightarrow$ xml_task$^*$
\end{lstlisting}
%

@@

let task_map = swap(command_map,time_map) in
let full_task_map = 
  ins("<Task>") . del("\n\n") . task_map . ins("</Task>") 
in
let repeated_task_map = full_task_map . iterate(full_task_map)
let nonempty_task_map = ins("<Tasks>") . repeated_task_map . ins("</Tasks>")

@@

Using these combinators, we can build up a lens that synchronizes crontabs and
Scheduled Task files, shown in Figure~\ref{fig:example_lens}. This lens uses one
typing rule not previously mentioned: type equivalence. While
\lstinline{scheduled_job} is a equivalent to a regular expression with an
outermost disjunction, it is not syntactically equivalent to one. Without type
equivalence, the provided lens wouldn't actually be well-typed, but type
equivalence states that if a lens is typed between a pair of regular
expressions, it is also typed between an equivalent pair.

Note that, while the programmed lens should be quite simple, it was relatively
difficult to write. We had to provide a lot of scaffolding to map between the
desired formats, even though the idea was quite simple: align the times but
leave the commands unaligned. Furthermore, this is inherent to manually writing
these synchronization functions as a programmer must describe which components
of the strings are aligned with which, and which components are scaffolding.

\subsection{Synthesizing Symmetric Lenses}
To alleviate these difficulties, we wish to synthesize symmetric lenses from
types and examples. Given a pair of regular expressions and a set of
input-output examples, we want to synthesize a symmetric lens expression that is
typed between the regular expression pair, and satisfies the input/output
examples. In this example, we wish to synthesize a lens from \lstinline{crontab}
and \lstinline{scheduled_job}, using the strings in
Figure~\ref{fig:minimized-representations} as an input-output example for
\CreateR.

However, our simple symmetric lens combinators permit many well-typed programs
between two regular expressions. For example, one possible lens between
\lstinline{crontab} and \lstinline{scheduled_job} is the one in
Figure~\ref{fig:example_lens}, but the lens
%
\begin{lstlisting}
disc(crontab,scheduled_job,rep(crontab),rep(scheduled_job))
\end{lstlisting}
%
is also a valid lens (where \lstinline{rep($\Regex$)} is a string in the
language of $\Regex$). Furthermore, adding the synchronized files from
Figure~\ref{fig:minimized-representations} as an input-output example helps
little. The default strings given to \Disconnect merely need to be the strings
in the input-output examples, and the lens will still satisfy the specification.
While the disconnect lens is prohibited by two examples, or by a carefully
chosen input-output example on \PutR or \PutL, this problem continues presenting
itself for the sublenses, and is highlighted even more on larger formats. While
we could merely ask programmers to provide large, tight specifications for their
synthesis tasks, we would prefer the algorithm to choose the right lens
automatically. Instead of merely finding \emph{any} satisfying lens, we wish to
synthesize the \emph{best} satisfying lens.

But what is the best satisfying lens? We think the best satisfying lens is the
one that aligns as much data as is permitted by the type system and the
examples. More formally, we wish to minimize the expected number of bits
required to recover data in one format from a synchronized string in the other.
The expected number of bits is a well known concept in information theory: the
expected number of bits of a data source is the \emph{entropy} of that data
source.

The entropy of a data source is based on the probability distribution of that
data source. Given a set $S$, and a probability distribution over that set $P
\OfType S \rightarrow \Reals$, the entropy of the $(S,P)$ pair is $-\Sigma_{s\in
  S}P(s)\Log_2P(s)$. While we have ways of expressing sets of strings (with
regular expressions), we do not have probability distributions. In this work, we
express these probability distributions jointly with the language through
\emph{stochastic regular expressions}, regular expressions annotated with
probabilities. In this work, we guess a reasonable probability distribution from
the syntax, though if users manually provided stochastic regular expressions, or
large suite of data in the formats through which we could learn the annotations,
we could instead use that.

With sets and probability distributions, we can find the expected number of bits
required to recover one string from a synchronized string. This expected value
is the conditional entropy of the stochastic regular expression ($\Regex$) given
another stochastic regular expression ($\RegexAlt$), and a lens that shows which
strings are synchronized with which ($\Lens$). We aim to find the lens that
minimizes the sum of that entropy with the conditional entropy of $\RegexAlt$
given $\Regex$ and $\Lens$. Now that we have a notion of best, we need a way to
search for that best lens.

Our overall approach involves two communicating synthesizers: one (\RXSearch)
that proposes regular expression pairs, semantically but not syntactically
equivalent to the originals; one (\GreedySynth) that tries to find the best lens
within a class of lenses, characterized by such regular expression pairs. Given
two regular expressions, \GreedySynth can quickly search through the class of
lenses that can be typed between them, where all instances of type equivalence
involve no star equivalences. For example, consider the lens shown in
Figure~\ref{fig:example_lens}. This lens would \emph{not} be in the class
characterized by \lstinline{crontab} and \lstinline{scheduled_job}, as that
typing derivation requires unrolling \lstinline{xml_task$^*$} within
\lstinline{scheduled_job} into \lstinline{"" | (xml_task . xml_task$^*$)}.

How does \RXSearch propose stochastic regular expression pairs? We define
rewrite rules for stochastic regular expressions, and prove those rewrites
preserve language and probability distributions. By applying these rewrites, we
can generate new regular expression pairs, syntactically distinct but
semantically equivalent to the originals.

Critically for this work, \GreedySynth efficiently searches for one of the best
lenses in a class of lenses, characterized by a stochastic regular expression
pair. \GreedySynth works by searching through a smaller search space. First,
\GreedySynth converts the provided stochastic regular expression pair into a
\emph{stochastic DNF regular expression} pair. Where a stochastic regular
expression pair characterizes a given class, a stochastic DNF regular expression
pair is highly normalized -- it is the \emph{unique} pair that characterizes a
given class. Then, we use the syntax of the DNF regular expression pair to
greedily find a highly normalized version of a simple symmetric lens, a
\emph{symmetric DNF lens}. Finally, this symmetric DNF lens is converted into a
simple symmetric lens.

However, these communicating synthesizers pose a problem: unless we find a lens
with no cost, further search may find a better lens. While we may have found a
great lens between two regular expressions, a different regular expression pair
may provide an even better lens. There may be a great lens in the class
characterized by \lstinline{crontab} and \lstinline{scheduled_job}, but it is
not the the intended lens shown in Figure~\ref{fig:example_lens}. However, there
is a tension, we want to find a good lens, but we also want to give quick
feedback to the user. After finding the lens in Figure~\ref{fig:example_lens},
we dont want the algorithm to spend much longer searching for better lenses.

We address this by keeping running track of the best lens found thus far, and
eventually cutting the proposal-synthesis loop short. Before proposing new
types, the regular expression generator, \RXSearch, evaluates whether it should
continue proposing regular expression pairs, or whether it is content with the
current best. For a given regular expression pair, \RXSearch looks at the number
of rewrite rules performed on the pair, and identifies how many pairs with
involving the same number of rewrites may be proposed. If there are a large
number of pairs that can be proposed with that many rewrites, or the current
lens is very well synchronized, \RXSearch does not propose that pairs. In this
way, \RXSearch will continue searching for better lenses, but does not tackle
huge search spaces unless it needs to. In this way, the algorithm can provide
lenses quickly to the users, which users can refine with additional examples.
% end overview

@@

\subsection{\GreedySynth}
\label{subsec:greedy-synth}
The second synthesizer, \GreedySynth, actually generates the lenses. It does
this by first converting the regular expressions into \emph{stochastic DNF
  regular expressions}, intuitively stochastic regular expressions with all
disjunctions distributed over all concatenations. Next, a good DNF lens is
greedily found between these stochastic DNF regular expressions, using the
probability annotations on the DNF regular expressions to find the cost.
Finally, this DNF lens is converted into a simple symmetric lens.
